\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-08 chapters/covert_align.tex
All text
Proofread

The decoding methods described in \autoref{sec:stbf-struct} and
\autoref{sec:bttda} are rather general attempts at increasing \ac{erp} decoding accuracy.
Given the problem statement of this work, we are also interested in targeted
methods that would specifically advance the development of gaze-independent
visuall \acp{bci} (see \label{sec:gaze-independence/sota}).
Methods of directly enhancing gaze-independent decoding through optimizing
decoder design have not been widely studied (see
\label{sec:gaze-independence/sota/decoding}).
We assume that advances in gaze-independent decoding can be made if properties
of covert \ac{vsa} that affect performance can be identified and, in some way,
accounted for.
The latency estimation and alignment approach proposed in \autoref{sec:wcble}
was specifically designed for this purpose.

\textcite{Arico2014} observed higher variability in single-trial P3 peak
latencies relative to stimulus onset during covert \ac{vsa} compared to overt
\ac{vsa}.
This latency variability contributes to reduced covert VSA decoding performance.
While they proposed an analysis and performance prediction method, they did not provide a
decoding solution.
They suggested that compensating for latency jitter could enhance covert VSA
decoding, but did not actually verify this hypothesis directly.
Additionally, \cite{Hardiansyah2020} developed a classifier for
covert VSA ERPs, exploiting single-trial latency features in combination with
amplitude features for classification with a support vector machine.
They demonstrated the positive influence of single-trial ERP component latency
features on covert VSA inference, yet did not attempt to correct the amplitude
features for these latencies.

Furthermore, as mentioned in~\autoref{sec:gaze-independence/sota/visual} and
\autoref{sec:gaze-independence/approach/cvsa-erp},
our goal is also to explore more flexible gaze-independent settings that the
usual covert \ac{vsa} with central gaze fixation reported in most literature.
\textcite{Frenzel2011} introduced a similar protocol to the proposed split
\ac{vsa} setting (see~\autoref{fig:gaze/vsa/free}).
They showed that it is possible to perform split VSA and that, in this
case, visuospatial attention and gaze direction can separately be decoded using
classical ERP techniques.
To the best of our
knowledge, this is the sole study that investigated split attention in ERP-based BCIs.
However, \textcite{Frenzel2011} considered their interface only for the case
where the user actively intends to select both targets determined by the gaze
and the VSA.
In the split VSA setting considered in our work, we rather instruct the participant to ignore
the distractor and only attend the cued target, since we are interested in
decoding the visuospatial attention only.

@@

\subsection{Data collection}
We evaluate our approach on a dataset specifically recorded for this study, designed to
probe different modalities of covert VSA, and on the publicly available
BNCI2014-009 dataset~\cite{Aloise2012a}.

@@

We recorded a dataset to validate our approach.
The Covert Visuospatial Attention ERP (CVSA-ERP) dataset
consists of 15 participants, mean age $26.34\pm3.04$ years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
Each subject performed different VSA conditions (overt, covert and split),
illustrated in Figure~\ref{fig:interface} (top row).
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \cite{Treder2010}, we present six flashing targets (without letters or
symbols) to the participant while the EEG, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The VSA conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:interface}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
	\end{subfigure}

@@

  \caption[Stimulation interfaces and evoked \acp{erp}.]{
		\subref{fig:interface} Interfaces and visuospatial attention (VSA)
    conditions in the CVSA-ERP and BNCI2014-009 datasets.
		In the CVSA-ERP oddball BCI interface, screen targets are intensified one after
    the other in pseudorandom order while the participant can either pay overt,
    covert, or split VSA to the cued target.
		In the BNCI2014-009 overt VSA interface, entire rows and columns are
		intensified at once. In its covert counterpart,
		groups of 6 letters are intensified one after the other, partly relying on
    feature attention.
		\subref{fig:erps} Contrast between target (color) and non-target, and
    distractor (gray) and non-target grand average event-related potentials per VSA condition and dataset.
    Overt VSA yields a strong modulation of the N1 component in both datasets;
    the P3 amplitude decreases with the degree of split VSA.
    In split VSA, N1 and P2 are more prominently evoked by the distractor,
    while the P3 is evoked by the target.
	}
\end{figure*}

In contrast to the protocol proposed by \cite{Frenzel2011}, split VSA was
performed by instructing the participant to attend the intensifications of the
cued target, and ignore the intensifications of the distractor target.
Since we assume there will be an effect depending on the distance between
attended target and the distractor, we discern three split VSA sub-conditions:
the distractor is either clockwise or counterclockwise directly next to the
attended target ($d=1$), there is one other target between the attended target and
the distractor ($d=2$), or the distractor is opposite to the intended target
($d=3$).

EEG for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier
(Compumedics Neuroscan, Australia) at 2048Hz and 62 Ag/AgCl active electrodes arranged in the
international 10-10 layout fitted to a standard electrode cap (EASYCAP GmbH,
Germany), with electrodes located at AFz and FCz as ground and reference respectively.
Using electrolyte gel, electrode impedances were brought below 5k$\Omega$.
Electrodes TP9 and TP10, used for off-line re-referencing, were directly
attached to the skin using stickers for better contact.
The power line frequency in Belgium is 50 Hz.
Participant's eye gaze was registered using an EyeLink 1000 Plus eye tracker (SR Research,
Canada) in non-fixation mode.

Participants signed the informed consent form and were seated at a distance of
60 cm before a CRT-emulating monitor (VPixx
Technologies, Canada) operating at a refresh rate of 120Hz, displaying 6
circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
with a radius of 12.28° of visual angle centered on the midpoint of the screen,
conforming to the interface proposed by \textcite{Treder2010}
(\autoref{fig:covert-align/interface/interface})
A hexagonal layout interface with an empty center and a low number of targets
counteracts target crowding and, as long as the subject’s gaze is within the hexagon of
targets, no other target can be between the subject’s gaze and a covertly
attended target.
Targets are full-contrast white and were intensified by scaling them to a
larger size (5.60° of visual angle,
(\autoref{fig:covert-align/interface/intensification}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
peripheral visual field.
Stimuli were presented using PsychopPy (version 2023.1.3)~\cite{Peirce2019}.

@@

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_3.tikz.tex}
    \caption[Gaze fixated on a target.]{In the overt and split \ac{vsa}
    settings, the
    participant is cued to fixate on one of the targets.}
    \label{fig:covert-align/interface/overt}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_4.tikz.tex}
    \caption[Gaze fixated centrally]{In the covert \ac{vsa} setting, the participant is
    cued to fixate the center of the interface.}
    \label{fig:covert-align/interface/covert}
  \end{subfigure}
\end{figure}

The participant was instructed to press the space bar when ready for a block
of stimulations.
Then, one target was indicated as the cue and the participant
was instructed to count the number of intensifications of the cued target
during the following block of stimulations.
After pressing the space bar again, a blue crosshair appeared, and the subject
was instructed to fixate their gaze on the blue crosshair for the duration of
the stimulation block (\autoref{fig:covert-align/interface/overt} and
\autoref{fig:covert-align/interface/overt}).
The position of this crosshair determined the VSA condition for this trial:
overt VSA when the crosshair was at the same location as the cued target,
covert VSA when the crosshair appeared in the center of the screen, and split
VSA when the crosshair appeared on a different target than the cued one.
After pressing the space bar again and a delay of 5 seconds, the stimulation block
starts.

All targets were intensified for a duration of 100 ms, in pseudorandom order.
The inter-stimulus-interval (ISI), the time between the onsets of subsequent intensifications,
was variable and consisted of a fixed 300ms interval (of which 100ms with an intensified target onscreen)
with 200ms uniform jitter added, resulting in an ISI
between 200 and 400 ms.
\Acp{isi} were jittered to counteract steady-state effects and residue in averaging. A
longer \ac{isi} will increase component amplitude and aid in counteracting temporal
autocorrelation for a higher statistical test precision.
In a block of stimulations, each target was intensified a pseudorandom number of times between 10 and 15.
This led to stimulation blocks with an average duration of 26.25 seconds. After a block of stimulations, an
input prompt appeared to enter the mentally counted number of intensifications.
After inputting this number, the subject was allowed to pause until pressing the space bar again.
In total, six blocks were presented for overt VSA, six blocks for covert VSA, 12 blocks for
split ($d=1$) VSA, 12 blocks for split ($d=2$) VSA and 6 blocks for split
($d=3$) VSA, covering all possible combinations of VSA conditions, cued targets and
crosshair locations.

The experiment started with five non-recorded practice stimulation blocks, one for
each of the 5 VSA conditions.
During these practice blocks, the participant received feedback about their gaze
position and counting accuracy.
Counting the instructions and the participant's response to the
input prompts, a block lasted about 30
seconds. In sum, the experiment featured approximately 45 minutes of
stimulation time.
After blocks 14 and 28, the participant was allowed to take a longer break.
Including these longer breaks, the experiment lasted approximately one hour.

\subsubsection{BNCI2014-009 dataset}
The
BNCI2014-009\footnote{\url{https://bnci-horizon-2020.eu/database/data-sets}}
dataset~\cite{Aloise2012a} was used in the analysis performed
in \cite{Arico2014}.
It contains data from 10 subjects (median age $24.5\pm1.9$
years)  that performed two spelling tasks illustrated in the second row of
Figure~\ref{fig:interface}: using the P3 Matrix speller
interface to exploit overt VSA, and the GeoSpell covert VSA interface.
To use the GeoSpell interface, the participant gazes at the fixation point at the
center of the screen, while groups of characters flash simultaneously in a
circular layout around the fixation point.
The user directs their visuospatial attention to the location where the intended letter is expected
to appear, and when it does, a P3 ERP component is expected to be evoked.
This results in a specific setting where both visuospatial attention and
feature attention (the attended letter) are exploited.
For a detailed description of the paradigm and dataset, we refer
to \cite{Aloise2012a}.

\subsection{Data processing and analysis}
\subsubsection{Preprocessing}
Analysis was performed using Python and the MNE software package (version
1.3.1)~\cite{Gramfort2013}.
All datasets were band-pass filtered between 0.1Hz and 20Hz with a 4th-order Butterworth filter.
Bad channels in the data were automatically detected using the RANSAC
method~\cite{Fischler1981} and rejected.
The recorded EEG was re-referenced
offline to the average of the mastoid electrodes TP9 and TP10.
Next, the EEG signals were corrected for eye movement artifacts using an
Independent Component Analysis (ICA).
Since we have access to EOG data for the CVSA-ERP dataset components correlating
significantly with the EOG were rejected.
For the BNCI2014-009 dataset, ICA components were manually rejected.
Finally, the EEG signal is divided into epochs ranging from 100ms before stimulus onset to 700ms after stimulus onset and down-sampled to 128Hz.
In both datasets, only 16 channels were kept for
analysis (Fz, FCz, Cz, CPz, Pz, Oz, F3, F4, C3, C4, CP3, CP4, P3, P4, PO7 and
PO8).

\subsubsection{Decoders}
We will compare \ac{wcble} and \ac{cble} as described in
\autoref{sec:wcble/methods}, with their base classifier and tLDA and with
Riemannian Geometry
approaches that rely on spatial covariance as features.
Together with tLDA, Riemannian Geometry generally reaches state-of-the-art decoding
performance~\cite{Lotte2018}.
We implemented two Riemannian Geometry pipelines.
The first one estimates shrunk covariances from the ERPs filtered with 6 XDAWN
filters, projects these covariances to a tangent space and classifies the
result using $L_2$-regularized logistic regression
(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
Secondly, we adopt the pipeline from \cite{Aydarkhanov2020}, since their work shows
favorable performance in the presence of single-trial ERP latency jitter.
Shrunk spatial covariance matrices are estimated from epochs that are
augmented by concatenating the average target and average non-target ERP as
extra channels, projected to tangent space and classified using
$L_2$-regularized logistic regression (ERPCov-TS-LR).

To evaluate performance, 6-fold cross-validation without shuffling was performed for both
datasets.
At each fold, classifiers were trained on five target selection blocks (300
epochs) and tested on one block (60 epochs) without overlap for CVSA-ERP.
For each subject and run in the BNCI2014-009 dataset, classifiers
were trained on five symbol selections (480 epochs) and tested on one symbol
selection (96 epochs) without overlap at each fold.
A window ranging from 0ms to 600ms after stimulus onset was used for CBLE and WCBLE.
With epochs ranging from -100ms to 700ms relative to stimulus onset, this
allows for extracting latencies ranging from -100ms to +100ms.

@@

Figure~\ref{fig:single_trial_roc_auc_diff} shows a comparison of areas under the
Receiver Operating Characteristic curve (ROC-AUC) for all pairs of tLDA, CBLE
and WCBLE for single-trial classification to investigate the contributions of
CBLE and WCBLE relative to their first-stage classifier tLDA.
For this evaluation, epochs were rejected when the peak-to-peak amplitude
exceeded 800$\mu V$ and, for the CVSA-ERP dataset, if the user's gaze differed more than
10 degrees of visual angle from the fixation crosshair.
\begin{figure}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:single_trial_roc_auc_diff}
		\includegraphics[width=\linewidth]{figures/covert_align/figure4a.pdf}
	\end{subfigure}

@@

	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:block_evaluation}
		\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
	\end{subfigure}
  \caption[Decoder performance.]{%
		\subref{fig:single_trial_roc_auc_diff} Difference in cross-validated
    single-trial classification receiver operating characteristic curve ($\Delta$ROC-AUC)
    between Classifier-based Latency Estimation (CBLE), Woody
    CBLE (WCBLE),	and their first-stage classifier (tLDA).
    95\% confidence intervals were determined using $k=1000$ bootstrapping.
    Our proposed WCBLE decoder outperforms tLDA and CBLE for covert and split
		($d=3$)	visuospatial attention (VSA). CBLE scores on par with tLDA.
		\subref{fig:block_evaluation} Cross-validated target selection accuracy for
    all decoders plotted n function of the number of test repetitions in
    different VSA conditions. Significance was determined using one-sided
    (WCBLE $>$ other) Wilcoxon rank-sum tests using False Discovery Rate correction
    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). WCBLE generally achieves highest
    covert VSA target selection accuracy.}
\end{figure}
Wilcoxon signed-rank tests controlled for multiple comparisons by Benjamini
and Hochberg's False Discovery Rate procedure (FDR) revealed that
for the BNCI2014-009 dataset, WCBLE significantly outperformed tLDA
($\Delta\mathrm{ROC-AUC} = 0.019$, $p=0.004$) and CBLE
($\Delta\mathrm{ROC-AUC} = 0.016$, $p=0.036$) for covert VSA
but was significantly outperformed by tLDA in overt VSA decoding
($\Delta\mathrm{ROC-AUC}=-0.004$, $p=0.040$).
For the CVSA-ERP dataset, WCBLE
also achieved significantly better covert VSA performance than tLDA ($\Delta\mathrm{ROC-AUC}
	= 0.023$, $p=0.041$) and CBLE ($\Delta\mathrm{ROC-AUC}
	= 0.036$, $p=0.024$) .
We found no significant difference in WCBLE performance over tLDA in the split
VSA conditions in the CVSA-ERP dataset, but results show a clear trend of
increase in WCBLE performance over tLDA and CBLE as $d$ increases
CBLE failed to significantly outperform its first-stage classifier tLDA in all
evaluated VSA conditions.
Table~\ref{tab:covert-align/results/single-trial} reports all single-trial classification scores for all considered models, datasets
and conditions.

\begin{table}
  \centering
  \makebox[\textwidth][c]{%
    \input{chapters/covert_align_performance.tex}
  }
  \caption{Cross-validated single-trial classification area under the receiver operating characteristic
curve for all evaluated models, visuospatial attention conditions and
  datasets.}
  \label{tab:covert-align/results/single-trial}
\end{table}

>>>>

The decoding methods described in \cref{sec:stbf-struct} and
\cref{sec:bttda} are general attempts at increasing \ac{erp} decoding accuracy.
Given the problem statement of this work, we are also interested in targeted
methods that would specifically advance the development of gaze-independent
visual \acp{bci} (see \cref{sec:gaze-independence/sota}).
Methods of directly enhancing gaze-independent decoding through optimizing
decoder design have not been widely studied (see
\cref{sec:gaze-independence/sota/decoding}).
We assume that advances in gaze-independent decoding can be made if properties
of covert \ac{vsa} that affect performance can be identified and accounted for.
The latency estimation and alignment approach proposed in \cref{sec:wcble}
was specifically designed for this purpose.

\textcite{Arico2014} observed higher variability in single-trial P3 peak
latencies relative to stimulus onset during covert \ac{vsa} compared to overt
\ac{vsa}.
This latency variability contributes to reduced covert \ac{vsa} decoding performance.
While they proposed an analysis and performance prediction method, they did not provide a
decoding solution.
They suggested that compensating for latency jitter could enhance covert \ac{vsa}
decoding, but did not verify this hypothesis directly.
Additionally, \cite{Hardiansyah2020} developed a classifier for
covert \ac{vsa} \acp{erp}, exploiting single-trial latency features in combination with
amplitude features for classification with a support vector machine.
They demonstrated the positive influence of single-trial \ac{erp} component latency
features on covert \ac{vsa} inference, yet did not attempt to correct the amplitude
features for these latencies.

Furthermore, as mentioned in \cref{sec:gaze-independence/sota/visual} and
\cref{sec:gaze-independence/approach/cvsa-erp},
our goal is also to explore more flexible gaze-independent settings than the
usual covert \ac{vsa} with central gaze fixation reported in most literature.
\textcite{Frenzel2011} introduced a similar protocol to the proposed split
\ac{vsa} setting (see~\cref{fig:gaze/vsa/free}).
They showed that it is possible to perform split \ac{vsa} and that, in this
case, visuospatial attention and gaze direction can be decoded separately using
classical \ac{erp} techniques.
To the best of our
knowledge, this is the sole study that investigated split attention in \ac{erp}-based \acp{bci}.
However, \textcite{Frenzel2011} considered their interface only for the case
where the user actively intends to select both targets determined by the gaze
and the \ac{vsa}.
In the split \ac{vsa} setting considered in our work, we instruct the participant to ignore
the distractor and only attend to the cued target, as we are interested in
decoding the visuospatial attention only.

@@

\subsection{Data collection}
We evaluate our approach on a dataset specifically recorded for this study, designed to
probe different modalities of covert \ac{vsa}, and on the publicly available
BNCI2014-009 dataset~\cite{Aloise2012a}.

@@

We recorded a dataset to validate our approach.
The Covert Visuospatial Attention \ac{erp} (CVSA-ERP) dataset
consists of 15 participants, mean age $26.34\pm3.04$ years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
Each subject performed different \ac{vsa} conditions (overt, covert, and split),
illustrated in Figure~\ref{fig:interface} (top row).
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \cite{Treder2010}, we presented six flashing targets (without letters or
symbols) to the participant while the \ac{eeg}, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The \ac{vsa} conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:interface}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
	\end{subfigure}

@@

  \caption[Stimulation interfaces and evoked \acsp{erp}.]{
		\subref{fig:interface} Interfaces and visuospatial attention (\ac{vsa})
    conditions in the CVSA-ERP and BNCI2014-009 datasets.
		In the CVSA-ERP oddball \ac{bci} interface, screen targets are intensified one after
    the other in pseudorandom order while the participant can either pay overt,
    covert, or split \ac{vsa} to the cued target.
		In the BNCI2014-009 overt \ac{vsa} interface, entire rows and columns are
		intensified at once. In its covert counterpart,
		groups of 6 letters are intensified one after the other, partly relying on
    feature attention.
		\subref{fig:erps} Contrast between target (color) and non-target, and
    distractor (gray) and non-target grand average event-related potentials per \ac{vsa} condition and dataset.
    Overt \ac{vsa} yields a strong modulation of the N1 component in both datasets;
    the P3 amplitude decreases with the degree of split \ac{vsa}.
    In split \ac{vsa}, N1 and P2 are more prominently evoked by the distractor,
    while the P3 is evoked by the target.
	}
\end{figure*}

In contrast to the protocol proposed by \cite{Frenzel2011}, split \ac{vsa} was
performed by instructing the participant to attend to the intensifications of the
cued target and ignore the intensifications of the distractor target.
Since we assume there will be an effect depending on the distance between
the attended target and the distractor, we discern three split \ac{vsa} sub-conditions:
the distractor is either clockwise or counterclockwise directly next to the
attended target ($d=1$), there is one other target between the attended target and
the distractor ($d=2$), or the distractor is opposite the intended target
($d=3$).

\ac{eeg} for the CVSA-ERP dataset was recorded using a SynAmps RT amplifier
(Compumedics Neuroscan, Australia) at 2048Hz and 62 Ag/AgCl active electrodes arranged in the
international 10-10 layout fitted to a standard electrode cap (EASYCAP GmbH,
Germany), with electrodes located at AFz and FCz as ground and reference, respectively.
Using electrolyte gel, electrode impedances were brought below 5k$\Omega$.
Electrodes TP9 and TP10, used for off-line re-referencing, were directly
attached to the skin using stickers for better contact.
The power line frequency in Belgium is 50 Hz.
The participant's eye gaze was registered using an EyeLink 1000 Plus eye tracker (SR Research,
Canada) in non-fixation mode.

Participants signed an informed consent form and were seated at a distance of
60 cm from a CRT-emulating monitor (VPixx
Technologies, Canada) operating at a refresh rate of 120Hz, displaying 6
circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
with a radius of 12.28° of visual angle centered on the midpoint of the screen,
conforming to the interface proposed by \textcite{Treder2010}
(\cref{fig:covert-align/interface/interface}).
A hexagonal layout interface with an empty center and a low number of targets
counteracts target crowding, and as long as the subject’s gaze is within the hexagon of
targets, no other target can be between the subject’s gaze and a covertly
attended target.
Targets are full-contrast white and were intensified by scaling them to a
larger size (5.60° of visual angle,
(\cref{fig:covert-align/interface/intensification}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
peripheral visual field.
Stimuli were presented using PsychoPy (version 2023.1.3)~\cite{Peirce2019}.

@@

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_3.tikz.tex}
    \caption[Gaze fixated on a target.]{In the overt and split \ac{vsa}
    settings, the
    participant is cued to fixate on one of the targets.}
    \label{fig:covert-align/interface/overt}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/covert_align/interface_4.tikz.tex}
    \caption[Gaze fixated centrally.]{In the covert \ac{vsa} setting, the participant is
    cued to fixate at the center of the interface.}
    \label{fig:covert-align/interface/covert}
  \end{subfigure}
\end{figure}

The participant was instructed to press the space bar when ready for a block
of stimulations.
Then, one target was indicated as the cue, and the participant
was instructed to count the number of intensifications of the cued target
during the following block of stimulations.
After pressing the space bar again, a blue crosshair appeared, and the subject
was instructed to fixate their gaze on the blue crosshair for the duration of
the stimulation block (\cref{fig:covert-align/interface/overt} and
\cref{fig:covert-align/interface/covert}).
The position of this crosshair determined the \ac{vsa} condition for this trial:
overt \ac{vsa} when the crosshair was at the same location as the cued target,
covert \ac{vsa} when the crosshair appeared in the center of the screen, and split
\ac{vsa} when the crosshair appeared on a different target than the cued one.
After pressing the space bar again and a delay of 5 seconds, the stimulation block
started.

All targets were intensified for a duration of 100 ms in pseudorandom order.
The inter-stimulus-interval (\ac{isi}), the time between the onsets of subsequent intensifications,
was variable and consisted of a fixed 300 ms interval (of which 100 ms with an intensified target onscreen)
with 200 ms of uniform jitter added, resulting in an \ac{isi}
between 200 and 400 ms.
\Acp{isi} were jittered to counteract steady-state effects and residue in averaging. A
longer \ac{isi} will increase component amplitude and aid in counteracting temporal
autocorrelation for higher statistical test precision.
In a block of stimulations, each target was intensified a pseudorandom number of times between 10 and 15.
This led to stimulation blocks with an average duration of 26.25 seconds. After a block of stimulations, an
input prompt appeared to enter the mentally counted number of intensifications.
After inputting this number, the subject was allowed to pause until pressing the space bar again.
In total, six blocks were presented for overt \ac{vsa}, six blocks for covert \ac{vsa}, 12 blocks for
split ($d=1$) \ac{vsa}, 12 blocks for split ($d=2$) \ac{vsa}, and 6 blocks for split
($d=3$) \ac{vsa}, covering all possible combinations of \ac{vsa} conditions, cued targets, and
crosshair locations.

The experiment started with five non-recorded practice stimulation blocks, one for
each of the five \ac{vsa} conditions.
During these practice blocks, the participant received feedback about their gaze
position and counting accuracy.
Counting the instructions and the participant's response to the
input prompts, a block lasted about 30 seconds. In sum, the experiment featured approximately 45 minutes of
stimulation time.
After blocks 14 and 28, the participant was allowed to take a longer break.
Including these longer breaks, the experiment lasted approximately one hour.

\subsubsection{BNCI2014-009 dataset}
The BNCI2014-009\footnote{\url{https://bnci-horizon-2020.eu/database/data-sets}}
dataset~\cite{Aloise2012a} was used in the analysis performed
by \cite{Arico2014}.
It contains data from 10 subjects (median age $24.5\pm1.9$
years) who performed two spelling tasks illustrated in the second row of
\cref{fig:interface}: using the P3 Matrix speller
interface to exploit overt \ac{vsa}, and the GeoSpell covert \ac{vsa} interface.
To use the GeoSpell interface, the participant gazes at the fixation point at the
center of the screen, while groups of characters flash simultaneously in a
circular layout around the fixation point.
The user directs their visuospatial attention to the location where the intended letter is expected
to appear, and when it does, a P3 \ac{erp} component is expected to be evoked.
This results in a specific setting where both visuospatial attention and
feature attention (the attended letter) are exploited.
For a detailed description of the paradigm and dataset, we refer
to \cite{Aloise2012a}.

\subsection{Data processing and analysis}
\subsubsection{Preprocessing}
Analysis was performed using Python and the MNE software package (version
1.3.1)~\cite{Gramfort2013}.
All datasets were band-pass filtered between 0.1 Hz and 20 Hz with a 4th-order Butterworth filter.
Bad channels in the data were automatically detected using the RANSAC
method~\cite{Fischler1981} and rejected.
The recorded \ac{eeg} was re-referenced
offline to the average of the mastoid electrodes TP9 and TP10.
Next, the \ac{eeg} signals were corrected for eye movement artifacts using
Independent Component Analysis (ICA).
Since we have access to \ac{eog} data for the CVSA-ERP dataset, components correlating
significantly with the \ac{eog} were rejected.
For the BNCI2014-009 dataset, ICA components were manually rejected.
Finally, the \ac{eeg} signal was divided into epochs ranging from 100 ms before stimulus onset to 700 ms after stimulus onset and down-sampled to 128 Hz.
In both datasets, only 16 channels were kept for
analysis (Fz, FCz, Cz, CPz, Pz, Oz, F3, F4, C3, C4, CP3, CP4, P3, P4, PO7, and
PO8).

\subsubsection{Decoders}
\label{sec:covert-align/methods/decoders}
We compared \ac{wcble} and \ac{cble} as described in
\cref{sec:wcble/methods}, with their base classifier \ac{tlda} and with
Riemannian Geometry approaches that rely on spatial covariance as features.
Together with \ac{tlda}, Riemannian Geometry generally achieves state-of-the-art decoding
performance~\cite{Lotte2018}.
We implemented two Riemannian Geometry pipelines.
The first estimates shrunk covariances from the \acp{erp} filtered with 6 XDAWN
filters, projects these covariances to a tangent space, and classifies the
result using $L_2$-regularized logistic regression
(XDAWNCov-TS-LR)~\cite{Cecotti2017}.
Secondly, we adopted the pipeline from \cite{Aydarkhanov2020}, as their work shows
favorable performance in the presence of single-trial \ac{erp} latency jitter.
Shrunk spatial covariance matrices were estimated from epochs that were
augmented by concatenating the average target and average non-target \ac{erp} as
extra channels, projected to tangent space, and classified using
$L_2$-regularized logistic regression (ERPCov-TS-LR).

To evaluate performance, 6-fold cross-validation without shuffling was performed for both
datasets.
At each fold, classifiers were trained on five target selection blocks (300
epochs) and tested on one block (60 epochs) without overlap for CVSA-ERP.
For each subject and run in the BNCI2014-009 dataset, classifiers
were trained on five symbol selections (480 epochs) and tested on one symbol
selection (96 epochs) without overlap at each fold.
A window ranging from 0 ms to 600 ms after stimulus onset was used for \ac{cble} and \ac{wcble}.
With epochs ranging from -100 ms to 700 ms relative to stimulus onset, this
allows for extracting latencies ranging from -100 ms to +100 ms.

@@

\Cref{fig:single_trial_roc_auc_diff} shows a comparison of areas under the
Receiver Operating Characteristic curve (ROC-AUC) for all pairs of \ac{tlda}, \ac{cble},
and \ac{wcble} for single-trial classification to investigate the contributions of
\ac{cble} and \ac{wcble} relative to their first-stage classifier \ac{tlda}.
For this evaluation, epochs were rejected when the peak-to-peak amplitude
exceeded 800$\mu V$, and for the CVSA-ERP dataset, if the user's gaze differed by more than
10 degrees of visual angle from the fixation crosshair.
\begin{figure}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:single_trial_roc_auc_diff}
		\includegraphics[width=\linewidth]{figures/covert_align/figure4a.pdf}
	\end{subfigure}

@@

	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:block_evaluation}
		\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
	\end{subfigure}
  \caption[Decoder performance.]{%
		\subref{fig:single_trial_roc_auc_diff} Difference in cross-validated
    single-trial classification receiver operating characteristic curve ($\Delta$ROC-AUC)
    between Classifier-based Latency Estimation (CBLE), Woody
    CBLE (WCBLE),	and their first-stage classifier (tLDA).
    95\% confidence intervals were determined using $k=1000$ bootstrapping.
    Our proposed WCBLE decoder outperforms tLDA and CBLE for covert and split
		($d=3$)	visuospatial attention (VSA). CBLE scores on par with tLDA.
		\subref{fig:block_evaluation} Cross-validated target selection accuracy for
    all decoders plotted as a function of the number of test repetitions in
    different VSA conditions. Significance was determined using one-sided
    (WCBLE $>$ other) Wilcoxon rank-sum tests using False Discovery Rate correction
    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). WCBLE generally achieves the highest
    covert VSA target selection accuracy.}
\end{figure}
Wilcoxon signed-rank tests controlled for multiple comparisons by Benjamini
and Hochberg's False Discovery Rate procedure (FDR) revealed that
for the BNCI2014-009 dataset, \ac{wcble} significantly outperformed \ac{tlda}
($\Delta\mathrm{ROC-AUC} = 0.019$, $p=0.004$) and \ac{cble}
($\Delta\mathrm{ROC-AUC} = 0.016$, $p=0.036$) for covert \ac{vsa},
but was significantly outperformed by \ac{tlda} in overt \ac{vsa} decoding
($\Delta\mathrm{ROC-AUC}=-0.004$, $p=0.040$).
For the CVSA-ERP dataset, \ac{wcble}
also achieved significantly better covert \ac{vsa} performance than \ac{tlda} ($\Delta\mathrm{ROC-AUC}
	= 0.023$, $p=0.041$) and \ac{cble} ($\Delta\mathrm{ROC-AUC}
	= 0.036$, $p=0.024$).
We found no significant difference in \ac{wcble} performance over \ac{tlda} in the split
\ac{vsa} conditions in the CVSA-ERP dataset, but results show a clear trend of
increased \ac{wcble} performance over \ac{tlda} and \ac{cble} as $d$ increases.
\ac{cble} failed to significantly outperform its first-stage classifier \ac{tlda} in all
evaluated \ac{vsa} conditions.
\Cref{tab:covert-align/results/single-trial} reports all single-trial classification scores for all considered models, datasets,
and conditions.

\begin{table}
  \centering
  \makebox[\textwidth][c]{%
    \input{chapters/covert_align_performance.tex}
  }
  \caption{Cross-validated single-trial classification area under the receiver operating characteristic
curve for all evaluated models, visuospatial attention conditions, and
  datasets.}
  \label{tab:covert-align/results/single-trial}
\end{table}

\Cref{fig:block_evaluation} shows the cross-validated BCI selection
accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
decoders.
Accuracy was determined by, for each block, selecting the character with the
highest (stage-two if applicable) classifier score and comparing it to the cued target.
Significance was calculated using one-sided Wilcoxon rank-sum tests ($p=0.05$)
corrected for FDR over decoders.
For this evaluation, no epochs were rejected to keep the trial-based structure of BCI
operation intact.
For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
Yet, \ac{wcble} yields an improved decoding accuracy for covert \ac{vsa} in both
datasets, which is greatest for smaller numbers of repetitions and
decreases as the number of repetitions increases.
This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in thdeep q learning tutoriale BNCI2014-009 dataset
for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
Furthermore, while we reported a relative decrease in single-trial ROC-AUC for
\ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
selection accuracy.
No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
While Riemannian methods are significantly outperformed by \ac{tlda}, \ac{cble}, and \ac{wcble}
in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
\ac{wcble} in CVSA-ERP.
